---
layout: post
title: mlr loves OpenML
author: Heidi Seibold
draft: true
---

```{r, include=FALSE}
library("knitr")
opts_chunk$set(eval=FALSE)
```


[OpenML](http://www.openml.org/) stands for Open Machine Learning and is an online platform, which aims
at supporting collaborative machine learning online. It is an Open Science project that allows its users
to share data, code and machine learning experiments.

At the time of writing this blog we are in Eindoven at an [OpenML workshop](http://openml2016dev.openml.org/), where developers and
scientists meet to work on improving the project. Some of these people are R users and we are
developing an [R package](https://github.com/openml/openml-r) (which is called OpenML) that connects to and communicates with the OpenML platform.

## OpenML in R
The OpenML R package can for example list and download data sets and machine learning tasks (prediction challenges).
In R then one can run algorithms and then upload the results to OpenML. The website then shows how well the algorithm performs.
To run the algorithm on a given task the OpenML R package builds on the [mlr](https://github.com/mlr-org/mlr) package. mlr understands what a task is and
can run learners on that task. So all the OpenML package needs to do is convert the OpenML objects to objects mlr 
understands and then mlr deals with the learning.


## A small case study

We want to create a little study on the [OpenML websit](http://www.openml.org/), in which we compare different types of Support Vector Machines.
The study gets an ID assigned to it, which in our case is 25.
For this we use the function ksvm from package kernlab, which is itegrated in mlr ("classif.ksvm").


For details on installing and setting up please see the [guide](https://github.com/openml/openml-r)
on GitHub.

Let's start conducting the study:

- Load the packages and list data sets and tasks which have between 100 and 500 
observations. Then merge the infos on data sets and tasks.
```{r, eval=TRUE, message=FALSE}
library("OpenML")
library("mlr")
library("farff")
```

```{r}
dsize = c(100, 500)
dchars = listOMLDataSets(number.of.instances = dsize)
tchars = listOMLTasks(number.of.instances = dsize)

mergeby = intersect(names(tchars), names(dchars))
mergeby = mergeby[mergeby != "tags"]
chars_all = merge(dchars, tchars, by = mergeby)
```

- Select all supervised classification tasks the do 10-fold cross-validation and choose only one task per
data set. To make it fast, we focus on only the first three tasks.
```{r}
chars_10cv = subset(chars_all, task.type == "Supervised Classification"
                    & estimation.procedure == "10-fold Crossvalidation"
                    & is.na(evaluation.measures))
dups = duplicated(chars_10cv$data.id)
chars = chars_10cv[!dups, ]
chars = chars[1:3, ]
```

- Create the learners we want to compare.
```{r}
lrn.list = list(
  makeLearner("classif.ksvm", type = "C-svc"),
  makeLearner("classif.ksvm", type = "kbb-svc"),
  makeLearner("classif.ksvm", type = "spoc-svc")
)
```

- Run the learners on the three tasks.
```{r}
grid = expand.grid(task.id = chars$task.id, 
                   lrn.ind = seq_along(lrn.list))

runs = lapply(seq_row(grid), function(i) {
  message(i)
  task = getOMLTask(grid$task.id[i])
  ind = grid$lrn.ind[i]
  runTaskMlr(task, lrn.list[[ind]])
})
```

- And finally upload the runs to OpenML. 
The upload function (uploadOMLRun) returns the ID of the uploaded run object.
When uploading runs that are part of a certain study, tag it with study_ and the
study ID.
```{r}
run.id = lapply(runs, uploadOMLRun, tags = "study_25")
```

```{r, eval=TRUE}
evals = listOMLRunEvaluations(tag = "study_25")

evals$task.id = as.factor(evals$task.id)
evals$setup.id = as.factor(evals$setup.id)

library("ggplot2")
ggplot(evals, aes(x = setup.id, y = predictive.accuracy, 
                  color = data.name, group = task.id)) + 
  geom_point() + geom_line()

```


## Further infos
If you are interested in more, check out the OpenML [blog](https://medium.com/open-machine-learning),
the [paper](https://www.researchgate.net/publication/263890323_OpenML_Networked_science_in_machine_learning)
and the [GitHub repos](https://github.com/openml).





