---
layout: post
title: "Paper is out; mlr: Machine Learning in R"
author: jakob
---

We are happy to announce that our paper is out and finally we can answer your question on how to cite *mlr* in your publication.

The paper is published in the Journal of Machine Learning Research (JMLR) and is open access. 
You can find the abstract, the bib-file and the PDF following this [link](http://www.jmlr.org/papers/v17/15-066.html).

<!--more-->

It offers a brief overview of the features of *mlr* and a short comparision to similar toolkits.
For an in-depth understanding we still recommend our [excellent mlr tutorial](https://mlr-org.github.io/mlr-tutorial/) which is also available as a [PDF on arxiv.org](https://arxiv.org/abs/1609.06146) - but due to technical limitations missing parts of the appendix. 

For offline enjoyment you can always download the [latest version of the tutorial zipped](https://mlr-org.github.io/mlr-tutorial/devel/mlr_tutorial.zip).

Once *mlr 2.10* is on cran you will find citation information calling:
```{r, eval=FALSE}
citation("mlr")
```
```{r, echo=FALSE}
tmp = capture.output(citation("mlr"))
tmp = stringi::stri_replace_all_fixed(tmp, "{{", "{% raw %}{{{% endraw %}")
cat(paste0(tmp, "\n"))
```

