---
layout: post
title: First release of mlrMBO - the toolbox for (Bayesian) Black-Box Optimization
author: jakob
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(cache = TRUE, collapse = FALSE)
knitr::knit_hooks$set(document = function(x){
  gsub("```\n*```r*\n*", "", x)
})
```

We are happy to be able to finally announce the first release of [**mlrMBO** on cran](https://cran.r-project.org/package=mlrMBO) after a quite long development time.
For the theoretical background and a nearly complete overview of mlrMBOs capabilities you can check our [paper on **mlrMBO** that we presubmitted to arxiv](https://arxiv.org/abs/1703.03373).

The key features of **mlrMBO** are:

* Global optimization of expensive Black-Box functions.
* Mulit-Criteria Optimization.
* Parallelization through multi-point proposals.
* Support for optimization over categorical variables using random forests as a surrogate.

For examples covering different scenarios we have Vignettes that are also available as an (online documentation)[https://mlr-org.github.io/mlrMBO/].
For **mlr** users **mlrMBO** is especially interesting for hyperparameter optimization.

<!-- more -->

**mlrMBO** for **mlr** hyperparameter tuning was already used in [an earlier blog post](How-to-win-a-drone-in-20-lines-of-R-code). 
Nonetheless we want to provide a small toy example to demonstrate the work flow of **mlrMBO** in this post.

### Example

First we define an objective function that we are going to minimize:

```{r objectiveFunction}
set.seed(1)
library(mlrMBO)
fun = makeSingleObjectiveFunction(
  name = "SineMixture",
  fn = function(x) sin(x[1])*cos(x[2])/2 + 0.04 * sum(x^2),
  par.set = makeNumericParamSet(id = "x", len = 2, lower = -5, upper = 5)
)
```

To define the objective function we use `makeSingleObjectiveFunction` from the neat package [**smoof**](https://github.com/jakobbossek/smoof), which gives us the benefit amongst others to be able to directly visualize the function.
_If you happen to be in need of functions to optimize and benchmark your optimization algorithm I recommend you to have a look at it!_

```{r plotObjectiveFunction}
library(ggplot2)
plot3D(fun, contour = TRUE, lightning = TRUE)
```

Let's start with the configuration of the optimization:

```{r makeMBOControl}
# In this simple example we construct the control object with the defaults:
ctrl = makeMBOControl()
# For this numeric optimization we are going to use the Expected Improvement as Infill Criterion
ctrl = setMBOControlInfill(ctrl, crit = crit.ei)
# We will allow for exactly 20 evaluations of the objective Function:
ctrl = setMBOControlTermination(ctrl, max.evals = 25L)
```

The optimization has to so start with an initial design.
**mlrMBO** can automatically create one but here we are going to use a randomly sampled LHS design our own:

```{r design}
des = generateDesign(n = 8L, par.set = getParamSet(fun), fun = lhs::randomLHS)
autoplot(fun, render.levels = TRUE) + geom_point(data = des)
```

The dots demonstrate how the initial design already covers the search space but is missing the global minimum.
Before we can start the Bayesian optimization we have to set the surrogate learner to *Kriging*.

```{r surrogate}
sur.lrn = makeLearner("regr.km", predict.type = "se", config = list(show.learner.output = FALSE))
```

_Note:_ **mlrMBO** can automatically determine a good surrogate learner based on the search space defined for the objective function.
For a purely numeric domain it would have chosen *Kriging* as well with some slight modifications to make it a bit more stable against numerical problems that can occur during optimization.

Finally, we can start the optimization run:

```{r mboRun}
res = mbo(fun = fun, design = des, learner = sur.lrn, control = ctrl, show.info = TRUE)
res$x
res$y
```

We can see that we have found the global optimum of $y = -0.414964$ at $x = (-1.35265,0)$ quite sufficiently.
Let's have a look at the points mlrMBO evaluated.
Therefore we can use the `OptPath` which stores all information about all evaluations within the optimization run:

```{r mboPoints}
library(plotly)
opdf = as.data.frame(res$opt.path)
opdf$label = with(opdf, sprintf("y:%.2f, pred.y:%.2f, iter:%i", y, mean, dob))
g = autoplot(fun, render.levels = TRUE, render.contours = FALSE) + geom_point(data = opdf, aes(text = label))
ggplotly(g)
```

### Comparison

That is all good, but how do other optimization strategies perform?

#### Grid Search

Grid search is seldom a good idea.
But especially for hyperparameter tuning it is still used.
Probably because it kind of gives you the feeling that you know what is going on and have not left out any important area of the search space.
In reality the grid is usually so sparse that it leaves important areas untouched as you can see in this example:

```{r gridSeach}
grid.des = generateGridDesign(par.set = getParamSet(fun), resolution = 5)
grid.des$y = apply(grid.des, 1, fun)
grid.des[which.min(grid.des$y),]
autoplot(fun, render.levels = TRUE, render.contours = FALSE) + geom_point(data = grid.des)
```

It is no surprise, that the grid search could not cover the search space well enough and we only reach a bad result.

#### What about a genetic algorithm?

```{r cmaes}
library(cmaesr)
cmaes(fun, control = list(stop.ons = c(list(stopOnMaxEvals(25L)), getDefaultStoppingConditions())))
```

Can also not convincing for this example.
Their advantage lays in higher dimensional spaces with cheaper evaluations.

#### What about a simple random search?

```{r randomSearch}
random.des = generateRandomDesign(par.set = getParamSet(fun), n = 25L)
random.des$y = apply(random.des, 1, fun)
random.des[which.min(random.des$y),]
autoplot(fun, render.levels = TRUE, render.contours = FALSE) + geom_point(data = random.des)
```

With the random search you could always be lucky but in average the optimum is not reached if smarter optimization strategies work well.
To be totally fair for the comparison we would have to repeat all stochastic optimization several times, like it was done [our **mlrMBO** paper](https://arxiv.org/abs/1703.03373).
The examples here are just meant to demonstrate the package.

### Engage

If you want to contribute to **mlrMBO** we ware always open to suggestions and pull requests on github.
You are also invited to fork the repository and build and extend your own optimizer based on our toolbox.




